---
title: 'S&DS 363 Problem Set 4: MANOVA & GLM'
author: "Franklin Wu and Teresa Nguyen"
date: "2025-03-04"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Human Development Index MANOVA / GLM

MANOVA is a way to analyze the group to predict differences in the multivariate means of continuous variables and in our study, we will be using MANOVA to examine if groups differ across of a combination of dependent variables to understand the relationship between the variables. 

### Data Cleaning and Useful Packages

```{r}
library(MASS)
library(biotools)
library(heplots)
library(klaR)
library(car)

```

Always need to start with this first. Read in Data, choose the columns that best fit our needs, and rename the columns.

```{r}
data <- read.csv("../../Documents/sds-363/Human Development Index - Full.csv")

# Based on going through the Kaggle dataset and choosing the most relevant features
data <- data[, c(
  "Country",
  "UNDP.Developing.Regions",
  "HDI.Rank..2021.",
  "Human.Development.Index..2021.",
  "Life.Expectancy.at.Birth..2021.",
  "Mean.Years.of.Schooling..2021.",
  "Gross.National.Income.Per.Capita..2021.",
  "Gender.Development.Index..2021.",
  "Coefficient.of.human.inequality..2021.",
  "Overall.loss......2021.",
  "Gender.Inequality.Index..2021.",
  "Maternal.Mortality.Ratio..deaths.per.100.000.live.births...2021.",
  "Adolescent.Birth.Rate..births.per.1.000.women.ages.15.19...2021.",
  "Labour.force.participation.rate..female....ages.15.and.older...2021.",
  "Labour.force.participation.rate..male....ages.15.and.older...2021.",
  "Carbon.dioxide.emissions.per.capita..production...tonnes...2021."
)]

# rename for better readability
colnames(data) <- c(
  "Country",
  "Region",
  "HDI_Rank_2021",
  "HDI_2021",
  "Life_Expectancy_2021",
  "Mean_Years_Schooling_2021",
  "GNI_Per_Capita_2021",
  "Gender_Dev_Index_2021",
  "Human_Inequality_Coeff_2021",
  "Overall_Loss_2021",
  "Gender_Inequality_Index_2021",
  "Maternal_Mortality_Rate_2021",
  "Adolescent_Birth_Rate_2021",
  "Female_Labour_Force_2021",
  "Male_Labour_Force_2021",
  "CO2_Emissions_percapita_2021"
)

# removes all rows in the data object that contain any missing (NA) values
# our data frame should only include complete cases 
data <- data[complete.cases(data) & data$Region != "", ]
data <- subset(data, Region != "ECA")
names(data)
data$Income_Level <- cut(data$GNI_Per_Capita_2021, breaks = c(-Inf, 4465, 13845, Inf), labels = c("Low-Middle", "Upper-Middle", "High"))
table(data$Region, data$Income_Level)

```

### Interaction Plots for Each Response Variable

We chose our categorical variables to be the region and income level and our response variables are life expectancy, human inequality coefficient, and mean years of schooling. 

```{r}

interaction.plot(data$Region, data$Income_Level, data$Life_Expectancy_2021, lwd = 3, col = c("red", "blue"), trace.label = "Income Level", xlab = "Region", ylab = "Life Expectancy", main = "Interaction Plot for Region, Income Level, and Life Expectancy")

interaction.plot(data$Region, data$Income_Level, data$Human_Inequality_Coeff_2021, lwd = 3, col = c("red", "blue"), trace.label = "Income Level", xlab = "Region", ylab = "Human Inequality Coefficient", main = "Interaction Plot for Region, Income Level, and Inequality Coef.")

interaction.plot(data$Region, data$Income_Level, data$Mean_Years_Schooling_2021, lwd = 3, col = c("red", "blue"), trace.label = "Income Level", xlab = "Region", ylab = "Mean Years of Schooling", main = "Interaction Plot for Region, Income Level, and Mean Years Schooling")
```

The interaction plot for *Life Expectancy* in relation to Region and Income Level suggests a possible interaction between the variables. The lines representing *Lower-Middle* and *Upper-Middle* income levels are not parallel, particularly as they approach the lower end of the life expectancy scale, around 60. The divergence possibly indicates that the effect of Region on Life Expectancy is not consistent across different income levels. Additionally, the *High-Income group* follows a different trajectory compared to the *Upper-Middle Income* group, suggesting greater variation in life expectancy across regions. If the lines were parallel, it would no interactions of Region and Life Expectancy across income levels. However, the lack of parallels between each of the income level groups seem to show that the impact of Region on Life Expectancy varies depending on Income Level, so there may be an interaction.

For the second plot, which examines the *Human Inequality Coefficient*, the *Lower-Middle* and *High-Income level* lines appear to be parallel, indicating no interaction. However, the *Lower-Middle* and *Upper-Middle* income lines converge and follow a different trajectory, suggesting a possible interaction between these income levels.

Finally, in the third plot analyzing Mean Years of Schooling, there is a possible interaction between the *High-Income* and *Upper-Middle Income groups*, as their lives are no parallel, showing **possibly** that Income Level significantly affects schooling outcomes in these categories; on the other hand, the Upper-Middle and Lower-Middle income groups appear to have parallel lines, indicating little to no interaction. 

### Two Way MANOVA

We hypothesize that mean vectors change according to the independent variables region and income level. 

```{r}
options(contrasts = c("contr.sum", "contr.poly"))
h_model <- lm(cbind(Life_Expectancy_2021, Human_Inequality_Coeff_2021, Mean_Years_Schooling_2021) ~ Region*Income_Level,  
                data = data)
summary(Anova(h_model, type = 3), univariate = T)
```

*Multivariate Results*: For each of the tests (Wilks, Pillalli, Hotelling, and Roy), region, income level, and interaction are all significant as all the p-values of each of these are under our critical value of 0.05 and that there is an interaction effect between region and income level. 

*Univaraite Results*: 
Income levels seem to strongly affect life expectancy with a p value of less than 0.05 (9.0349e-08), suggesting that higher income levels are associated with longer life expectancies. Regions have a strong effect on life expectancy, but not as strong as income levels with a p-value of 0.0026, which may be because the life expectancy is different among regions. There is a strong interaction however, between region, income level, and life expectancy, which shows that income and life expectancy varies by region. 
There seems to be a that all the variables, region, income level, and the interactino between region and income level seem to greatly affect the human inequality coefficient, showing that inequality levels vary by region and income levels. 
For mean years of schooling, region and the interaction between income and region are not significant with p values greater than 0.05. However, the income level is significant with p = 6.2479e-08 < 0.05, suggesting that schooling levels increase with income. 

### Multivariate and Univariate Contrasts

```{r}

linearHypothesis(h_model, "Income_Level1 + Income_Level2 = 0", test = "Wilks")

levels(data$Income_Level)
colnames(coef(h_model))

```

*Multivariate contrast tests*

In our first pair of test, we tested a contrast comparing the combined impact of the factors Middle Income and SSA vs Upper Income and SSA. Our first test gave us a p-value of 0.0017 (less than our alpha of 0.05), which suggests that the combined effect of Middle Income and Sub-Saharan Africa (SSA) is significantly different from zero. In contrast, our second test looking at the factors upper income and SSA had a p-value of 0.3585, which is larger than our p-value of 0.05 and means we do not have statistically significant evidence to reject the null hypothesis. The findings from these 2 multi-variate contrast tests suggest that the factors Middle Income and SSA (for households) have a strong combined effect on the outcomes (Life Expectancy, Human Inequality Coefficient, Mean Years of Schooling), while the combined effect of SSA and Upper Income do not have that effect. Based on some preliminary conjecture, it may be that middle income households in SSA exhibit distinct patterns in life expectancy, education, or inequality that make their combination highly influential while upper income households in SSA may 1) lack the same set of patterns relating to these outcomes or 2) it may be that our data has lesser households in the Upper Income threshold. 

In another multivariate contrast test we looked at whether being in Europe (and Central Asia) and High Income differed from being Middle Income in Sub-Saharan Africa. Our results gave us a p-value of 0.014, which means we can reject the null hypothesis that these combined effects are equal. This finding has many interesting implications as it means wealthier countries in Europe and Central Asia (which have higher amounts of wealth on average) impact our outcome variables differently in comparison to middle income countries in SSA. This likely addresses many underlying factors identified in economics research - for instance, higher income countries in ECA tend to have better infrastructure (which directly impacts our outcome variables relating to health and education) in combination with other factors such as stronger and more political institutions in higher income countries in ECA compared to more middle level income countries in SSA. 

*Univariate contrast tests*

We also took a look at univariate contrast tests in reference to the outcome variable of life expectancy. Our first test looked at whether Region 1 (Asia) had a significant impact on our outcome variable and our second test looked at whether Region 1 and Region 2 (EAP) have significantly different effects on life expectancy. Our results show that the p-value of 0.2115 and 0.6763 respectively are greater than the alpha of 0.05, which means we cannot reject the null hypothesis that these regions have a stastistically significant impact on the outcome variable of just life expectancy.

```{r}
#DONT eliminate outlier - messes up model
#data <- subset(data, CO2_Emissions_percapita_2021 < 7)

plot(data$CO2_Emissions_percapita_2021, data$Life_Expectancy_2021, 
     main = "CO2_Emissions_percapita_2021 vs. Life Expectancy", xlab = "CO2_Emissions_percapita_2021", ylab = "Life Expectancy", pch = 19)

plot(data$CO2_Emissions_percapita_2021, data$Human_Inequality_Coeff_2021, 
     main = "CO2_Emissions_percapita_2021 vs. Inequality Coeff", xlab = "CO2_Emissions_percapita_2021", ylab = "Inequality Coefficient", pch = 19)

plot(data$CO2_Emissions_percapita_2021, data$Mean_Years_Schooling_2021, 
     main = "CO2_Emissions_percapita_2021 vs. Years of Schooling", xlab = "CO2_Emissions_percapita_2021", ylab = "Mean Years Schooling", pch = 19)

h_model_cont <- lm(cbind(Life_Expectancy_2021, Human_Inequality_Coeff_2021, Mean_Years_Schooling_2021) ~ Region * Income_Level + CO2_Emissions_percapita_2021, data = data)

summary(Anova(h_model_cont, type = 3), univariate = TRUE)

```
```{r}
#my heplots not working
library(heplots)
cqplot(h_model_cont$residuals)

```
```{r}

# Load required package
library(vegan)

(mrpp_region <- mrpp(data[, c("Life_Expectancy_2021", "Human_Inequality_Coeff_2021", "Mean_Years_Schooling_2021")], data$Region,))


(mrpp_income <- mrpp(data[, c("Life_Expectancy_2021", "Human_Inequality_Coeff_2021", "Mean_Years_Schooling_2021")],data$Income_Level,))


```

